Do researchers preferentially collaborate with colleagues of the same gender?

### 2018-05-29 Gender_homophily_analysis.Rmd
Finalise markdown

---
title: "Do researchers preferentially collaborate with same-gendered colleagues?"
subtitle: Online Supplementary Material
author: "Luke Holman and Claire Morandin"
output:
  html_document: 
    toc: true # show a table of contents
    toc_float: true # make table of contents float at the side of the page
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  # set to true if you want to number sections at each table header
    theme: lumen # nice fonts and colours
    code_folding: hide # Add some buttons to show/hide the code
---

@@

```{r warning=FALSE, message=FALSE, results='hide'}
library(dplyr) # data handling 
library(stringr) # string manipulation
library(ggplot2) # plots
library(parallel) # parallel computing
library(lme4) # mixed models
library(lmerTest) # mixed models
library(pander) # nice Markdown tables
library(kableExtra) # scrollable html tables
library(tidyr) # spread and gather functions
library(gridExtra) # arranging ggplots
library(mgcv) # generalised additive models
library(ggbeeswarm) # beeswarm plots
library(Cairo) # for exporting the character 'alpha' via ggsave

# Using a database allows quick access to the whole PubMed dataset without loading the whole thing to memory.
pubmed_db <- src_sqlite("~/Desktop/Gender Pubmed mining project/data for analysis/pubmed_gender_data.sqlite3", create = F)
pubmed_sqlite <- tbl(pubmed_db, "papers") # information about each individual paper
journals_sqlite <- tbl(pubmed_db, "journals") # information about each journal
```

```{r echo=FALSE, results='hide'}
library(knitr) # set up cacheing to save time when re-building the html document - chunks are not re-evaluated if the code has not changed
opts_chunk$set(warning=FALSE, message=FALSE) # Suppress warnings in the HTML output

figure.counter <- 2  # Counters for numbering figures and tables
supp.figure.counter <- 1
table.counter <- 1
supp.table.counter <- 1
```
`r opts_chunk$set(cache=TRUE)`

@@

  # Only do the test if over.write = TRUE or there is no datafile in the working directory
  if(over.write | !(file.name %in% list.files())){

@@

### Define a function to calculate Bergstrom et al.'s 'coefficient of homophily', $\alpha$
These two functions estimate $\alpha$ for a set of papers' author lists. The version `find.alpha.CIs()` performs bootstrapping resampling of the dataset to obtain the 95% confidence limits on $\alpha$. `find.alpha()` calculates $\alpha$ without bootstrapping, making it much faster (used later when we are calculating the null distribution of $\alpha$).

@@

  # Only do the test if over.write = TRUE or there is no datafile in the working directory
  if(over.write | !(output.file %in% list.files())){ 
    journals <- names(count.data)
    nJournals <- length(count.data)
    # Set up a cluster
    cl <- makePSOCKcluster(number.of.cores)
    setDefaultCluster(cl)
    clusterExport(NULL, c("find.alpha",
                          "find.alpha.CIs", 
                          "find.alpha.specific.author",
                          "find.alpha.specific.author.CIs",
                          "gender.grouping.test.one.journal",
                          "pubmed_sqlite"))
    clusterEvalQ(NULL, library(dplyr))

@@

# Get all the data we need out of the database, and save as an RDS file - for papers in the last 12 months
retrieve.authorship.data(unique.journals, 
                         minimum.n = minimum.n, 
                         start.of.time.period = 1, 
                         end.of.time.period = 0, 
                         file.name = "last.year.rds", 
                         over.write = FALSE)

# Get all the data we need out of the database, and save as an RDS file - for papers that are between 10 and 11 years old
retrieve.authorship.data(unique.journals, 
                         minimum.n = minimum.n, 
                         start.of.time.period = 11, 
                         end.of.time.period = 10, 
                         file.name = "older.papers.rds", 
                         over.write = FALSE)

@@

  count.data <- readRDS(paste(filename, ".rds", sep = "")) 
  data.frame(Quantity = c("Number of disciplines", 
                          "Number of journals", 
                          "Number of papers",
                          "Number of authors",
                          "Median number of papers per journal", 
                          "Median number of authors per journal",
                          "Median number of authors per paper"),
             Value = c(journal.data %>% filter(journal %in% names(count.data)) %>% 
                         select(discipline) %>% 
                         distinct() %>% nrow(),
                       length(count.data),
                       nrow(do.call("rbind", count.data)[,1:2]),
                       sum(do.call("rbind", count.data)[,1:2]),
                       median(sapply(count.data, nrow)),
                       median(sapply(count.data, function(x) sum(x[,1]) + sum(x[,2]))),
                       median(rowSums(do.call("rbind", count.data)[,1:2])))) 
}

sample.size.table <- make.sample.size.table("older.papers") %>% 
  rename(`n (2005-2006)` = Value) %>% 
  left_join(make.sample.size.table("last.year") %>% 
              rename(`n (2015-2016)` = Value), by = "Quantity")
write.csv(sample.size.table, file = "manuscript/sample.size.table.csv", row.names = FALSE)

sample.size.table %>% pander(split.cell = 40, split.table = Inf)
```

@@

```{r get_alpha}
# Calculate alpha across all authorship positions for all suitable NEW papers
gender.grouping.test.many.journals(readRDS("last.year.rds"),
                                   "all",
                                   "homophily.results.all.csv",
                                   nBoots = 1000, 
                                   nSimulations = 1000, 
                                   number.of.cores = 8,
                                   over.write = FALSE)

# Calculate alpha for first authors for all suitable NEW papers
gender.grouping.test.many.journals(readRDS("last.year.rds"),
                                   "first",
                                   "homophily.results.first.authors.csv",
                                   nBoots = 1000, 
                                   nSimulations = 1000, 
                                   number.of.cores = 8,
                                   over.write = FALSE)

# Calculate alpha for last authors for all suitable NEW papers
gender.grouping.test.many.journals(readRDS("last.year.rds"),
                                   "last",
                                   "homophily.results.last.authors.csv",
                                   nBoots = 1000, 
                                   nSimulations = 1000, 
                                   number.of.cores = 8,
                                   over.write = FALSE)

# Calculate alpha across all authorship positions for all suitable OLD papers
gender.grouping.test.many.journals(readRDS("older.papers.rds"),
                                   "all",
                                   "older.homophily.results.all.csv",
                                   nBoots = 1000, 
                                   nSimulations = 1000, 
                                   number.of.cores = 8,
                                   over.write = FALSE)

# Calculate alpha for first authors for all suitable OLD papers
gender.grouping.test.many.journals(readRDS("older.papers.rds"),
                                   "first",
                                   "older.homophily.results.first.authors.csv",
                                   nBoots = 1000, 
                                   nSimulations = 1000, 
                                   number.of.cores = 8,
                                   over.write = FALSE)

# Calculate alpha for last authors for all suitable OLD papers
gender.grouping.test.many.journals(readRDS("older.papers.rds"),
                                   "last",
                                   "older.homophily.results.last.authors.csv",
                                   nBoots = 1000, 
                                   nSimulations = 1000, 
                                   number.of.cores = 8,
                                   over.write = FALSE)

@@

# Load up all the homophily results, and merge them with the journal-specific information
gender.homophily.all <- merge.and.clean("homophily.results.all.csv", journal.data)
gender.homophily.first <- merge.and.clean("homophily.results.first.authors.csv", journal.data)
gender.homophily.last <- merge.and.clean("homophily.results.last.authors.csv", journal.data)

gender.homophily.all.old <- merge.and.clean("older.homophily.results.all.csv", journal.data)
gender.homophily.first.old <- merge.and.clean("older.homophily.results.first.authors.csv", journal.data)
gender.homophily.last.old <- merge.and.clean("older.homophily.results.last.authors.csv", journal.data)

# Bind all 5 datasets into a single one, with a column called 'Subset' to identify them
master.dataset <- rbind(gender.homophily.all,
                        gender.homophily.first,
                        gender.homophily.last,
                        gender.homophily.all.old,
                        gender.homophily.first.old,
                        gender.homophily.last.old) %>%
  mutate(Subset = unlist(mapply(rep, c("All authors", "First authors", "Last authors"), 
                                each = c(nrow(gender.homophily.all), 
                                         nrow(gender.homophily.first),
                                         nrow(gender.homophily.last),
                                         nrow(gender.homophily.all.old),
                                         nrow(gender.homophily.first.old),
                                         nrow(gender.homophily.last.old)))),
         Age = unlist(mapply(rep, c("New", "Old"), 
                                each = c(sum(nrow(gender.homophily.all), 
                                             nrow(gender.homophily.first), 
                                             nrow(gender.homophily.last)),
                                         sum(nrow(gender.homophily.all.old), 
                                             nrow(gender.homophily.first.old), 
                                             nrow(gender.homophily.last.old)))))) %>%
  select(discipline, journal, Subset, Age, standardised.IF, n.useable.papers, n.authors, 
         gender.ratio.of.sample, observed.alpha, lower.95.CI, upper.95.CI, alpha.adjusted,
         lower.95.CI2, upper.95.CI2, two.tail.p.value, adjusted.p.value) %>%
  rename(unadjusted.p.value = two.tail.p.value)

# Save and export a neat and tidy version for the supplementary material
neat <- master.dataset
names(neat) <- gsub("[.]", " ", names(neat))
names(neat) <- gsub("Subset", "Author position", names(neat))
names(neat) <- paste(toupper(substring(names(neat), 1, 1)), substring(names(neat), 2), sep="")
names(neat) <- gsub("N", "n", names(neat))
names(neat) <- gsub("2", "", names(neat))
for(i in which(sapply(neat, is.numeric))) neat[,i] <- round(neat[,i], 3)
write.csv(neat, file = "Supplementary dataset S1.csv", row.names = FALSE)
rm(neat)
```

@@

if(!("country.data.csv" %in% list.files())){ # Write the data, assuming it's not already been done
  country.data <- gender.grouping.test.by.journal.and.country(country.data)
  write.csv(country.data, file = "country.data.csv", row.names = FALSE)
}

country.data <- read.csv("country.data.csv", stringsAsFactors = FALSE) %>%
  mutate(p.adjusted = p.adjust(two.tail.p.value, method = "BH")) %>%
  left_join(journal.data %>% select(journal, discipline), by = "journal") # %>% 
  # left_join(master.dataset %>% 
  #             filter(Age == "New" & Subset == "All authors") %>%
  #             select(journal, alpha.adjusted) %>%
  #             rename(journal.alpha = alpha.adjusted),
  #           by = "journal")
```

@@

@@

**Table S`r supp.table.counter`**: The table shows similar information to Figure S`r supp.figure.counter`, namely the distribution of the $\alpha'$ values across journals, split by the research discipline. The gender ratio column shows the percentage of women authors in the sample used to calculate $\alpha'$, across all authorship positions. In the last two columns, the numbers outside parentheses give the number of journals that deviate statistically significantly from zero, while the numbers inside parentheses give the number that remain significant after false discovery rate correction.

@@

```{r}
mean.inflation.due.to.country <- country.data %>%
  group_by(journal) %>%
  summarise(mean.inflation.due.to.country = mean(journal.alpha - alpha.adjusted))

@@

simulation.figure
ggsave(simulation.figure, file = "figures/figure5.pdf", 
       height = 5.6, width = 5)#, device = cairo_pdf)
```

>>>>

---
title: "Do researchers preferentially collaborate with colleagues of the same gender?"
subtitle: Online Supplementary Material
author: "Luke Holman and Claire Morandin"
output:
  html_document: 
    toc: true # show a table of contents
    toc_float: true # make table of contents float at the side of the page
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  # set to true if you want to number sections at each table header
    theme: lumen # nice fonts and colours
    code_folding: hide # Add some buttons to show/hide the code
---

@@

```{r warning=FALSE, message=FALSE, results='hide'}
library(dplyr) # data handling 
library(stringr) # string manipulation
library(ggplot2) # plots
library(parallel) # parallel computing
library(lme4) # mixed models
library(lmerTest) # mixed models
library(pander) # nice Markdown tables
library(tidyr) # spread and gather functions
library(gridExtra) # arranging ggplots
library(mgcv) # generalised additive models
library(ggbeeswarm) # beeswarm plots
library(Cairo) # for exporting the character 'alpha' via ggsave

# Using a database allows quick access to the whole PubMed dataset without loading the whole thing to memory.
# You can download the pubmed_gender_data.sqlite3 database here: https://osf.io/bt9ya/
pubmed_db <- src_sqlite("~/Desktop/Gender Pubmed mining project/data for analysis/pubmed_gender_data.sqlite3", create = F)
pubmed_sqlite <- tbl(pubmed_db, "papers") # information about each individual paper
journals_sqlite <- tbl(pubmed_db, "journals") # information about each journal
```

```{r echo=FALSE, results='hide'}
library(knitr) 
opts_chunk$set(warning=FALSE, message=FALSE) # Suppress warnings in the HTML output

figure.counter <- 2  # Counters for numbering figures and tables
supp.figure.counter <- 1
table.counter <- 1
supp.table.counter <- 1
```

@@

  # Only do the test if over.write = TRUE or there is no datafile in the working directory
  if(over.write | !file.exists(file.name)){

@@

### Define a function to calculate Bergstrom et al.'s "coefficient of homophily", $\alpha$
These two functions estimate $\alpha$ for a set of papers' author lists. The version `find.alpha.CIs()` performs bootstrapping resampling of the dataset to obtain the 95% confidence limits on $\alpha$. `find.alpha()` calculates $\alpha$ without bootstrapping, making it much faster (used later when we are calculating the null distribution of $\alpha$).

@@

  # Only do the test if over.write = TRUE or there is no datafile in the working directory
  if(over.write | !file.exists(output.file)){ 
    journals <- names(count.data)
    nJournals <- length(count.data)
    # Set up a cluster
    cl <- makePSOCKcluster(number.of.cores)
    setDefaultCluster(cl)
    clusterExport(NULL, c("find.alpha",
                          "find.alpha.CIs", 
                          "find.alpha.specific.author",
                          "find.alpha.specific.author.CIs",
                          "gender.grouping.test.one.journal",
                          "pubmed_sqlite"))
    clusterEvalQ(NULL, library(dplyr))

@@

# Get all the data we need out of the database, and save as an RDS file - for papers in the last 12 months
retrieve.authorship.data(unique.journals, 
                         minimum.n = minimum.n, 
                         start.of.time.period = 1, 
                         end.of.time.period = 0, 
                         file.name = "data/last.year.rds", 
                         over.write = FALSE)

# Get all the data we need out of the database, and save as an RDS file - for papers that are between 10 and 11 years old
retrieve.authorship.data(unique.journals, 
                         minimum.n = minimum.n, 
                         start.of.time.period = 11, 
                         end.of.time.period = 10, 
                         file.name = "data/older.papers.rds", 
                         over.write = FALSE)

@@

  count.data <- readRDS(paste("data/", filename, ".rds", sep = "")) 
  data.frame(Quantity = c("Number of disciplines", 
                          "Number of journals", 
                          "Number of papers",
                          "Number of authors",
                          "Median number of papers per journal", 
                          "Median number of authors per journal",
                          "Median number of authors per paper"),
             Value = c(journal.data %>% filter(journal %in% names(count.data)) %>% 
                         select(discipline) %>% 
                         distinct() %>% nrow(),
                       length(count.data),
                       nrow(do.call("rbind", count.data)[,1:2]),
                       sum(do.call("rbind", count.data)[,1:2]),
                       median(sapply(count.data, nrow)),
                       median(sapply(count.data, function(x) sum(x[,1]) + sum(x[,2]))),
                       median(rowSums(do.call("rbind", count.data)[,1:2])))) 
}

if(!file.exists("manuscript/sample.size.table.csv")){
  sample.size.table <- make.sample.size.table("older.papers") %>% 
    rename(`n (2005-2006)` = Value) %>% 
    left_join(make.sample.size.table("last.year") %>% 
                rename(`n (2015-2016)` = Value), by = "Quantity")
  write.csv(sample.size.table, file = "manuscript/sample.size.table.csv", row.names = FALSE)
}

read.csv("manuscript/sample.size.table.csv") %>% pander(split.cell = 40, split.table = Inf)
```

@@

```{r get_alpha}
# Calculate alpha across all authorship positions for all suitable NEW papers
gender.grouping.test.many.journals(readRDS("data/last.year.rds"),
                                   "all",
                                   "data/homophily.results.all.csv",
                                   nBoots = 1000, 
                                   nSimulations = 1000, 
                                   number.of.cores = 8,
                                   over.write = FALSE)

# Calculate alpha for first authors for all suitable NEW papers
gender.grouping.test.many.journals(readRDS("data/last.year.rds"),
                                   "first",
                                   "data/homophily.results.first.authors.csv",
                                   nBoots = 1000, 
                                   nSimulations = 1000, 
                                   number.of.cores = 8,
                                   over.write = FALSE)

# Calculate alpha for last authors for all suitable NEW papers
gender.grouping.test.many.journals(readRDS("data/last.year.rds"),
                                   "last",
                                   "data/homophily.results.last.authors.csv",
                                   nBoots = 1000, 
                                   nSimulations = 1000, 
                                   number.of.cores = 8,
                                   over.write = FALSE)

# Calculate alpha across all authorship positions for all suitable OLD papers
gender.grouping.test.many.journals(readRDS("data/older.papers.rds"),
                                   "all",
                                   "data/older.homophily.results.all.csv",
                                   nBoots = 1000, 
                                   nSimulations = 1000, 
                                   number.of.cores = 8,
                                   over.write = FALSE)

# Calculate alpha for first authors for all suitable OLD papers
gender.grouping.test.many.journals(readRDS("data/older.papers.rds"),
                                   "first",
                                   "data/older.homophily.results.first.authors.csv",
                                   nBoots = 1000, 
                                   nSimulations = 1000, 
                                   number.of.cores = 8,
                                   over.write = FALSE)

# Calculate alpha for last authors for all suitable OLD papers
gender.grouping.test.many.journals(readRDS("data/older.papers.rds"),
                                   "last",
                                   "data/older.homophily.results.last.authors.csv",
                                   nBoots = 1000, 
                                   nSimulations = 1000, 
                                   number.of.cores = 8,
                                   over.write = FALSE)

@@

# Load up all the homophily results, and merge them with the journal-specific information
gender.homophily.all <- merge.and.clean("data/homophily.results.all.csv", journal.data)
gender.homophily.first <- merge.and.clean("data/homophily.results.first.authors.csv", journal.data)
gender.homophily.last <- merge.and.clean("data/homophily.results.last.authors.csv", journal.data)

gender.homophily.all.old <- merge.and.clean("data/older.homophily.results.all.csv", journal.data)
gender.homophily.first.old <- merge.and.clean("data/older.homophily.results.first.authors.csv", journal.data)
gender.homophily.last.old <- merge.and.clean("data/older.homophily.results.last.authors.csv", journal.data)

# Bind all 5 datasets into a single one, with a column called 'Subset' to identify them
master.dataset <- rbind(gender.homophily.all,
                        gender.homophily.first,
                        gender.homophily.last,
                        gender.homophily.all.old,
                        gender.homophily.first.old,
                        gender.homophily.last.old) %>%
  mutate(Subset = unlist(mapply(rep, c("All authors", "First authors", "Last authors"), 
                                each = c(nrow(gender.homophily.all), 
                                         nrow(gender.homophily.first),
                                         nrow(gender.homophily.last),
                                         nrow(gender.homophily.all.old),
                                         nrow(gender.homophily.first.old),
                                         nrow(gender.homophily.last.old)))),
         Age = unlist(mapply(rep, c("New", "Old"), 
                                each = c(sum(nrow(gender.homophily.all), 
                                             nrow(gender.homophily.first), 
                                             nrow(gender.homophily.last)),
                                         sum(nrow(gender.homophily.all.old), 
                                             nrow(gender.homophily.first.old), 
                                             nrow(gender.homophily.last.old)))))) %>%
  select(discipline, journal, Subset, Age, standardised.IF, n.useable.papers, n.authors, 
         gender.ratio.of.sample, observed.alpha, lower.95.CI, upper.95.CI, alpha.adjusted,
         lower.95.CI2, upper.95.CI2, two.tail.p.value, adjusted.p.value) %>%
  rename(unadjusted.p.value = two.tail.p.value)
```

@@

if(!file.exists("data/country.data.csv")){ # Write the data, assuming it's not already been done
  country.data <- gender.grouping.test.by.journal.and.country()
  write.csv(country.data, file = "data/country.data.csv", row.names = FALSE)
}

country.data <- read.csv("data/country.data.csv", stringsAsFactors = FALSE) %>%
  mutate(p.adjusted = p.adjust(two.tail.p.value, method = "BH")) %>%
  left_join(journal.data %>% select(journal, discipline), by = "journal") 
```

@@

### Create S1 Data, which holds all the $\alpha'$ estimates

**S1 Data**: This spreadsheet shows the $\alpha'$ values calculated for each journal, in the 2005 and 2015 samples, and for each type of author (all authors, first authors, and last authors). The tables gives the impact factor of each journal, the sample size, $\alpha$ and $\alpha'$ and their 95% CIs, and the p-value from a 2-tailed test evaluating the null hypothesis that $\alpha$ is zero (both raw and FDR-corrected p-values are shown).

```{r}
# Save and export a neat and tidy version for the supplementary material
neat <- master.dataset
names(neat) <- gsub("[.]", " ", names(neat))
names(neat) <- gsub("Subset", "Author position", names(neat))
names(neat) <- paste(toupper(substring(names(neat), 1, 1)), substring(names(neat), 2), sep="")
names(neat) <- gsub("N", "n", names(neat))
names(neat) <- gsub("2", "", names(neat))
for(i in which(sapply(neat, is.numeric))) neat[,i] <- round(neat[,i], 3)
write.csv(neat, file = "data/S1 data.csv", row.names = FALSE)
rm(neat)
```

@@

**Table S`r supp.table.counter`**: The table shows similar information to Figure S`r figure.counter`, namely the distribution of the $\alpha'$ values across journals, split by the research discipline. The gender ratio column shows the percentage of women authors in the sample used to calculate $\alpha'$, across all authorship positions. In the last two columns, the numbers outside parentheses give the number of journals that deviate statistically significantly from zero, while the numbers inside parentheses give the number that remain significant after false discovery rate correction.

@@

```{r}
mean.inflation.due.to.country <- country.data %>%
  left_join(master.dataset %>% 
              filter(Age == "New", Subset == "All authors") %>% 
              select(journal, alpha.adjusted) %>% 
              rename(journal.alpha = alpha.adjusted), 
            by = "journal") %>%
  group_by(journal) %>%
  summarise(mean.inflation.due.to.country = mean(journal.alpha - alpha.adjusted))

@@

simulation.figure
ggsave(simulation.figure, file = "figures/figure5.pdf", 
       height = 5.6, width = 5, device = cairo_pdf)
```
